\chapter{Computational and Mathematical Techniques}
% If you like chapter abstracts ...
\dblspace
\begin{quote}{\em \input{Ch4/abstract4}}\end{quote}

\section{Introduction}
\label{sec:review:introduction}
CONFIRMATION REPORT
The bulk of the theory and technologies underpinning the research is expounded here.
\section{Segmentation}
  The methods of threshold segmentation, fast marching filtering and threshold segmentation level set filtering are explained, along with their advantages over other segmentation methods. The segmentation pipeline is delineated. A plan for this section is complete, and the writing needs to be expanded.
\section{Registration}
  The motivations and ideas behind registration are briefly discussed, and the generic structure of a registration pipeline is given. The individual components of the pipeline are examined, and the strengths and weaknesses of specific algorithms and strategies are presented in each case. Masking and sampling are touched upon. This section is effectively complete.
\section{Modelling Cardiac Electrophysiology}
We discuss cell ionic models and anisotropic diffusion based on tissue structure. The ionic model section needs significant work, but the tissue structure section was included in the transfer thesis.
\section{Finite Element Methods}
  The mathematical basis for the bidomain and monodomain equations is treated, along with their solution on a finite element mesh. I have a very strong grasp on the mathematics in both cases; the bidomain and monodomain section was included with the transfer, and the finite element section must be written up.
CONFIRMATION REPORT


