%!TEX root = ../thesis.tex
\vspace*{20mm}
{
\Large\bf
\begin{center}
Acknowledgements
\end{center}
}

This thesis is the culmination of many years' work and would not have been possible without the help and support of many people. 

I would like to thank my supervisors Dr. Martin Bishop, Dr. Blanca Rodriguez and Dr. Vicente Grau. From the very beginning, Martin has been the strongest mentor and a role model for achievement. Martin has provided tutoring, practical expertise, professional advice, friendship, opportunity, life coaching, and help with pretty much everything, unrecognisably beyond what could be expected, and greatly to my betterment. Blanca has imparted the vision, structure, guidance and scientific context, in my writing, project planning and management that I utterly lacked upon undertaking this endeavour. Her focus on direction, outcome and purpose has translated into many other parts of my life and has transformed my effectiveness as a person. Both Martin's and Blanca's commitment of pastoral care has been humbling and in embarrasing contrast to the experience of many of my peers; without it I would never have completed this course. In all aspects of the imaging side of this project, from theoretical and methodological creation and development, through complex debugging, to analysis and evaluation of results, Vicente has imparted unerring clarity and insight. His patient (and often repeated!) explanation and direction has beaten seemingly insurmountable problems and failures. All my supervisors have shown Olympian patience, generosity, support and care. I am a profoundly wiser and more able person having worked with them, and I look forward to returning their kindness in the future.

The thesis presented is in no small part been made possible through discussions with members of the Computational Biology Group and beyond. In particular, I owe a great deal to Raf Bordas, who has on several occasions proffered game-changing advice on technology and computational techniques, and who's cumulative saving of my wasted time is measured in months. In the same way, I would also like to thank Joe Pitt-Francis, Pras Pathmanathan, Jon Cooper, Miguel Bernabeu, Kevin Burrage, Mikael Wallman, Robbie Shade and Gabriel Rosser for thrillingly geeky and productive exchanges.

As always, my Mum and Dad have been just amazing over the past few years, and always there for support when I needed it most. I would especially like to thank my Mum for being so consistently enthused about my research career, and I would especially like to thank my Dad for consistently hiding his abject and heart-broken disappointment in my career choice. I am very excited to take them and my sister on holidays, excursions and general treats from now on!

My research was funded by the EPSRC, through the Systems Biology Doctoral Training Centre. I was given a solid educational foundation in research techniques by the DTC, and made great friends there, discussions with many of whom have contributed to this thesis.
